---
title: "From rel. abundance data to microbial community dynamics"
subtitle: "Computational 16S practical — EMBO Microbial Communities 2026"
author: "Fabian Springer"
date: "7 October 2026"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 9,
  fig.height = 5
)

suppressPackageStartupMessages({
  library(tidyverse)
  library(vegan)
  library(scales)
  library(ggrepel)
})

document_dir <- getwd()
project_root <- normalizePath(
  file.path(document_dir, ".."),
  mustWork = TRUE
)
document_path <- file.path(
  document_dir,
  "16S_analysis_and_visualization.Rmd"
)
abundance_path <- file.path(
  project_root,
  "data",
  "combined_relative_abundance.csv"
)
prepared_metadata_path <- file.path(project_root, "data", "metadata.tsv")

selected_rank <- "S"
top_n_taxa <- 10L
sum_tolerance <- 1e-6
```

# Learning objectives

By the end of this practical, you should be able to:

1. inspect and validate a taxonomic relative-abundance table;
2. reshape sparse taxonomic results into a complete sample-by-taxon matrix;
3. visualize community composition without forgetting that the data are compositional;
4. compare samples using Bray–Curtis dissimilarity and principal coordinate analysis (PCoA); and
5. connect barcode-level results to the experimental design using sample metadata.

> **Practical scope.** Taxonomic classification has already been run. We start from the combined NanoCLUST relative-abundance output so that the session can focus on interpretation rather than waiting for a sequencing pipeline.

Before loading the taxonomic profiles, import the cleaned experimental sample
sheet. Its `Sample_ID` column contains the barcode identifier used by the
NanoCLUST output. The identity of the six `Vincent_*` samples is unresolved, so
they are excluded from both the metadata and abundance profiles before any
summaries, plots, or distance calculations. They can be restored once their
experimental role has been clarified.

```{r import-experimental-metadata}
stopifnot(file.exists(prepared_metadata_path))

experimental_metadata_raw_df <- read_tsv(
  prepared_metadata_path,
  show_col_types = FALSE,
  progress = FALSE
)

stopifnot(
  all(
    c(
      "Sample_ID",
      "embl_id",
      "axenic_xenic_media",
      "timepoint_d",
      "replicate"
    ) %in% names(experimental_metadata_raw_df)
  ),
  !anyNA(experimental_metadata_raw_df$Sample_ID),
  !anyDuplicated(experimental_metadata_raw_df$Sample_ID)
)

sample_meta_df <- experimental_metadata_raw_df %>%
  filter(!str_detect(axenic_xenic_media, "^Vincent_")) %>%
  mutate(
    condition = axenic_xenic_media,
    condition = factor(condition, levels = c("Axenic", "Xenic", "Media")),
    timepoint = parse_number(timepoint_d),
    replicate = factor(replicate),
    sample_label = str_c(
      "T",
      timepoint,
      "_",
      str_to_lower(as.character(replicate))
    )
  ) %>%
  select(
    Sample_ID,
    sample_label,
    condition,
    timepoint,
    replicate,
    everything()
  )

stopifnot(
  nrow(experimental_metadata_raw_df) - nrow(sample_meta_df) == 6L,
  nrow(sample_meta_df) == 18L,
  !anyNA(
    sample_meta_df[
      c(
        "Sample_ID",
        "sample_label",
        "condition",
        "timepoint",
        "replicate"
      )
    ]
  ),
  !anyDuplicated(sample_meta_df$Sample_ID),
  !anyDuplicated(sample_meta_df[c("condition", "sample_label")])
)
```

# 1. Load and inspect the NanoCLUST output

The input contains one taxonomic assignment per sample and taxonomic rank. `tax` uses the codes `O`, `F`, `G`, and `S` for order, family, genus, and species.

```{r import-data}
stopifnot(file.exists(abundance_path))

abundance_raw_df <- read_csv(
  abundance_path,
  show_col_types = FALSE,
  progress = FALSE
)

required_columns <- c("sample", "tax", "taxid", "rel_abundance")
stopifnot(all(required_columns %in% names(abundance_raw_df)))

abundance_df <- abundance_raw_df %>%
  transmute(
    Sample_ID = sample %>%
      str_remove("^rel_abundance_") %>%
      str_remove("_concat$"),
    tax_rank = tax,
    taxon = taxid,
    rel_abundance = as.numeric(rel_abundance)
  ) %>%
  semi_join(
    sample_meta_df %>% select(Sample_ID),
    by = "Sample_ID"
  )

# get all available tax ranks
all_tax_ranks <- unique(abundance_df$tax_rank)

stopifnot(
  !anyNA(abundance_df),
  !anyDuplicated(abundance_df[c("Sample_ID", "tax_rank", "taxon")]),
  setequal(sample_meta_df$Sample_ID, abundance_df$Sample_ID),
  all(abundance_df$rel_abundance >= 0),
  all(abundance_df$rel_abundance <= 1)
)
```

Before plotting, summarize the object rather than printing all rows.

```{r compact-summary}
rank_labels <- c(O = "Order", F = "Family", G = "Genus", S = "Species")

input_summary_df <- abundance_df %>%
  group_by(tax_rank) %>%
  summarise(
    rank = unname(rank_labels[first(tax_rank)]),
    samples = n_distinct(Sample_ID),
    taxa = n_distinct(taxon),
    nonzero_assignments = n(),
    .groups = "drop"
  ) %>%
  select(rank, everything(), -tax_rank)

knitr::kable(input_summary_df)
```

## Check that each profile sums to one

Relative-abundance profiles should sum to one separately for every sample and rank.

```{r validate-profile-sums}
profile_sum_df <- abundance_df %>%
  group_by(Sample_ID, tax_rank) %>%
  summarise(profile_sum = sum(rel_abundance), .groups = "drop")

stopifnot(all(abs(profile_sum_df$profile_sum - 1) < sum_tolerance))

profile_sum_df %>%
  summarise(
    profiles_checked = n(),
    minimum_sum = min(profile_sum),
    maximum_sum = max(profile_sum)
  ) %>%
  knitr::kable(digits = 6)
```

**Discuss:** What information can we check in a relative-abundance table? What information has already been lost compared with the raw read counts?

# 2. Choose a taxonomic rank and complete missing combinations

We use species-level assignments for this practical because full-length Nanopore 16S reads can provide fine taxonomic resolution. Species labels are still hypotheses: classification depends on sequence quality, the reference database, conserved 16S regions, and the classifier.

The input is sparse: a taxon absent from a sample has no row. For plotting and distance calculations, we explicitly complete those missing sample–taxon combinations with zero.

```{r complete-species-table}
analysis_df <- abundance_df %>%
  filter(tax_rank == selected_rank) %>%
  select(Sample_ID, taxon, rel_abundance) %>%
  complete(Sample_ID, taxon, fill = list(rel_abundance = 0))

stopifnot(
  !anyNA(analysis_df),
  !anyDuplicated(analysis_df[c("Sample_ID", "taxon")])
)

completed_sum_df <- analysis_df %>%
  group_by(Sample_ID) %>%
  summarise(profile_sum = sum(rel_abundance), .groups = "drop")

stopifnot(all(abs(completed_sum_df$profile_sum - 1) < sum_tolerance))
```

# 3. Which organisms dominate the profiles?

A stacked bar chart gives an intuitive first view. To keep the legend readable, the ten most abundant taxa across all samples are shown individually and the remainder are grouped as `Other taxa`.

```{r prepare-composition}
taxon_summary_df <- analysis_df %>%
  group_by(taxon) %>%
  summarise(
    mean_abundance = mean(rel_abundance),
    maximum_abundance = max(rel_abundance),
    prevalence = mean(rel_abundance > 0),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_abundance))

top_taxa <- taxon_summary_df %>%
  slice_head(n = top_n_taxa) %>%
  pull(taxon)

sample_order <- sample_meta_df %>%
  arrange(condition, timepoint, replicate, sample_label) %>%
  pull(Sample_ID)

composition_df <- analysis_df %>%
  left_join(
    sample_meta_df %>% select(Sample_ID, sample_label, condition),
    by = "Sample_ID"
  ) %>%
  mutate(
    display_taxon = if_else(taxon %in% top_taxa, taxon, "Other taxa"),
    Sample_ID = factor(Sample_ID, levels = sample_order)
  ) %>%
  group_by(condition, Sample_ID, sample_label, display_taxon) %>%
  summarise(rel_abundance = sum(rel_abundance), .groups = "drop")

display_levels <- c(top_taxa, "Other taxa")
taxon_colors <- setNames(
  c(RColorBrewer::brewer.pal(length(top_taxa), "Paired"), "grey70"),
  display_levels
)

composition_df <- composition_df %>%
  mutate(display_taxon = factor(display_taxon, levels = rev(display_levels)))
```

```{r composition-plot}
composition_plot <- ggplot(
  composition_df,
  aes(x = Sample_ID, y = rel_abundance, fill = display_taxon)
) +
  geom_col(width = 0.9) +
  facet_grid(~condition, scales = "free_x", space = "free_x") +
  scale_fill_manual(values = taxon_colors, breaks = display_levels) +
  scale_x_discrete(
    labels = setNames(sample_meta_df$sample_label, sample_meta_df$Sample_ID)
  ) +
  scale_y_continuous(labels = label_percent(), expand = expansion(mult = c(0, 0.02))) +
  labs(
    title = "Species-level community profiles",
    subtitle = "Top taxa by mean relative abundance; remaining taxa are grouped",
    x = "Sample",
    y = "Relative abundance",
    fill = "Taxon"
  ) +
  theme_bw(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1),
    panel.grid.minor = element_blank()
  )

composition_plot
```

```{r dominant-taxa-summary}
dominant_taxon_df <- analysis_df %>%
  group_by(Sample_ID) %>%
  slice_max(rel_abundance, n = 1, with_ties = FALSE) %>%
  ungroup()

dominant_taxon_df %>%
  count(taxon, name = "samples_dominated", sort = TRUE) %>%
  knitr::kable()
```

**Discuss:** A highly abundant assignment may represent a bacterial community member, a host-associated organelle signal, contamination, or a classification artefact. What additional evidence would you want before interpreting it biologically?

## A compositionality thought experiment

Every sample sums to 100%. If one taxon doubles while all absolute cell counts remain unchanged, the relative abundance of every other taxon must fall. Therefore, relative abundance alone cannot tell us whether a taxon grew, declined, or merely changed relative to another organism.

# 4. Compare complete communities with Bray–Curtis dissimilarity

Bray–Curtis dissimilarity compares the complete abundance profiles of two samples. It is zero for identical profiles and approaches one as profiles become increasingly different.

```{r bray-curtis-pcoa}
species_mat <- analysis_df %>%
  pivot_wider(names_from = taxon, values_from = rel_abundance) %>%
  arrange(match(Sample_ID, sample_order)) %>%
  column_to_rownames("Sample_ID") %>%
  as.matrix()

stopifnot(
  is.numeric(species_mat),
  !anyNA(species_mat),
  all(abs(rowSums(species_mat) - 1) < sum_tolerance)
)

bray_distance <- vegdist(species_mat, method = "bray")
pcoa_fit <- cmdscale(bray_distance, k = 2, eig = TRUE, add = TRUE)

positive_eigenvalues <- pcoa_fit$eig[pcoa_fit$eig > 0]
axis_variance <- 100 * pcoa_fit$eig[1:2] / sum(positive_eigenvalues)

pcoa_df <- as_tibble(pcoa_fit$points, rownames = "Sample_ID") %>%
  rename(PCo1 = V1, PCo2 = V2) %>%
  left_join(
    sample_meta_df %>%
      select(Sample_ID, sample_label, condition),
    by = "Sample_ID"
  )

stopifnot(
  nrow(pcoa_df) == nrow(sample_meta_df),
  !anyNA(pcoa_df[c("Sample_ID", "sample_label", "condition")])
)

condition_levels <- levels(pcoa_df$condition)
condition_colors <- setNames(
  colorRampPalette(RColorBrewer::brewer.pal(8, "Dark2"))(
    length(condition_levels)
  ),
  condition_levels
)
```

```{r pcoa-plot}
pcoa_plot <- ggplot(pcoa_df, aes(PCo1, PCo2, color = condition)) +
  geom_point(size = 3) +
  geom_text_repel(
    aes(label = sample_label),
    size = 3,
    max.overlaps = 20,
    min.segment.length = 0,
    seed = 420
  ) +
  scale_color_manual(values = condition_colors) +
  labs(
    title = "Bray-Curtis PCoA of species profiles",
    subtitle = "Each point is one sample; labels show timepoint and replicate",
    x = sprintf("PCo1 (%.1f%%)", axis_variance[1]),
    y = sprintf("PCo2 (%.1f%%)", axis_variance[2]),
    color = "Condition"
  ) +
  coord_equal() +
  theme_bw(base_size = 13) +
  theme(panel.grid.minor = element_blank())

pcoa_plot
```

**Discuss:** Which samples are similar? Do the clusters reflect condition,
time, sequencing artefacts, or host-associated signal?

# 5. Follow abundant taxa through time

The retained experimental metadata map every axenic, xenic, and media sample
to its condition, sampling time in days, and biological replicate.

```{r taxon-time-courses}
time_course_taxa <- taxon_summary_df %>%
  slice_head(n = 6) %>%
  pull(taxon)

time_course_df <- analysis_df %>%
  filter(taxon %in% time_course_taxa) %>%
  left_join(sample_meta_df, by = "Sample_ID")

ggplot(
  time_course_df,
  aes(
    x = timepoint,
    y = rel_abundance,
    color = condition,
    group = interaction(condition, replicate)
  )
) +
  geom_line(alpha = 0.55) +
  geom_point(aes(shape = replicate), size = 2) +
  facet_wrap(~taxon, scales = "free_y") +
  scale_color_manual(values = condition_colors) +
  scale_y_continuous(labels = label_percent()) +
  labs(
    title = "Trajectories of the six most abundant assignments",
    x = "Timepoint (days)",
    y = "Relative abundance",
    color = "Condition",
    shape = "Replicate"
  ) +
  theme_bw(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "italic")
  )
```

**Discuss:** Do replicate cultures follow similar trajectories? Which changes appear condition-specific? What could relative abundance conceal?

# 6. What we can—and cannot—conclude

## Supported by this table

- which taxonomic assignments dominate each relative profile;
- which samples have similar overall species compositions;
- whether replicate trajectories appear qualitatively consistent once metadata are added; and
- which taxa are candidates for follow-up.

## Not supported without additional inputs

- sequencing depth or library-size quality control;
- rarefaction or count-based alpha diversity;
- absolute microbial growth or decline;
- a reliable xenic/axenic or time-course interpretation without barcode metadata; and
- confirmatory hypothesis testing without a prespecified repeated-measures design.

# Further steps (optional)

> **Draft extension.** These prompts provide a temporary starting point for
> participants who finish early. Trainers can refine or replace them as the
> practical develops.

## Compare taxonomic ranks

Change `selected_rank` near the beginning of the document from species (`S`)
to genus (`G`) and rerun the analysis.

- Which composition and ordination patterns remain stable?
- Which apparent differences depend on the selected taxonomic rank?
- Why might genus-level assignments be more reliable for some 16S reads?

## Investigate an unexpected dominant assignment

Choose one abundant assignment from the axenic samples and examine where it
occurs across conditions and timepoints.

- Is it specific to axenic cultures or also present in xenic and media samples?
- Could it represent a bacterium, host-associated chloroplast signal,
  contamination, or a database/classification limitation?
- What additional evidence would distinguish these explanations?

## Focus on one shared timepoint

Choose a timepoint represented in both axenic and xenic cultures and compare
the two biological replicates directly.

- Are within-condition replicates more similar than the two conditions?
- Which assignments contribute most strongly to the difference?
- Does the conclusion change when you inspect individual samples instead of
  condition averages?

# Take-home messages

1. Validate the input and experimental mapping before interpreting a plot.
2. Missing taxon rows are not missing samples; complete them explicitly with zero.
3. Relative abundances are constrained to sum to one and cannot establish absolute growth.
4. Ordination summarizes differences—it does not explain their biological cause.
5. Host-associated signal, database choice, and taxonomic uncertainty are part of the result, not nuisances to hide.

# Session information

```{r session-information}
sessionInfo()
```
